Certified Desugaring of Javascript Programs using Coq

### 2016-12-14 paper/paper.tex
working on the introduction

The JavaScript language is standardized by ECMA under the name
ECMAScript. Several versions of the standard now exist, the latest
ones being ECMAScript 5.1 (ES5, released in 2011), ECMAScript 2015, and
ECMAScript 2016.\footnote{The ECMAScript standard is now published yearly.} Its role is to make portability between different
JavaScript implementations possible by specifying the intended behavior
of the various language constructs and built-in functions of JavaScript.
The specification also makes it possible to develop formal semantics
for the language, which can then be used for, e.g., developing
static analysis tools.

One such semantics is \lambdajs. It aims to capture
the semantics of JavaScript by defining a simple core language
(based on the untyped call-by-value lambda calculus), and then
transforming (or ``desugaring'') the full JavaScript language into this
simple core. The \lambdajs semantics was originally developed to model the 
ECMAScript 3 specification~\cite{Guha-al:ECOOP10}, and it was later
revised~\cite{Politz-al:DLS12} to handle ECMAScript 5.1
features.\footnote{The authors named the revised language S5; however,
we continue to name it \lambdajs for simplicity.}
The interpreter for the core language and the JavaScript-to-core
transformation were developed using 
OCaml\cite{github:brownplt/LambdaS5};
the interpreter was tested using the official test262 test suite.
The \lambdajs semantics is useful for tool developers, as it
allows the tools to analyze JavaScript programs by working
on the simple core language of \lambdajs, having all the complexities
of JavaScript handled in the desugared code. It was already
successfully used to verify ADsafe~\cite{Politz-al:SEC11}.

More recently, JSCert~\cite{Bodin-al:POPL14},
formalized the ECMAScript 5.1 specification in the form of
a pretty-big-step semantics~\cite{Chargueraud:ESOP13} expressed in 
the form of a Coq inductive predicate. It aims to be as close to the
(informal English) specification as possible by maintaining close correspondence
between the evaluation rules and steps in the ES5 pseudocode. The project
also includes JSRef, a reference interpreter for JavaScript written in Coq
and proven correct with respect to JSCert. The interpreter is also tested
using the official ECMA test suite, test262. Thanks to the correctness proof,
testing of JSRef increases the confidence we can have that JSCert
correctly models ES5.

We believe is worthwhile to formally relate the two semantics. One reason is,
because the two semantics use a different approach, relating
them will increase the confidence we can place in both of them,
and is likely to find inconsistencies or other problems. Another reason is
that having \lambdajs formalized and certified (by proving correctness
with respects to JSCert) enables development of certified analysis tools
using \lambdajs as the core.

This is easier said than done. The task is very complex, mostly due
to the complexities of the JavaScript language. The ES5 specification,
upon which JSCert and \lambdajs are based, is 258 pages long and 
has 16 chapters. The JSCert pretty-big-step semantics, which models
the specification, has over 900 rules; the entire Coq source code
for the semantics has over 10000 lines. The \lambdajs environment
file, which contains the ECMAScript algorithms expressed in core
\lambdajs code, is almost 5000 lines long.

@@

\section{Formalization of \lambdajs}

@@

@@

\section{Correctness proof for \lambdajs desugaring}

@@

>>>>

The JavaScript language is standardized by ECMA under the name ECMAScript.
Several versions of the specification now exist, the latest ones being
ECMAScript 5.1 (ES5, released in 2011), ECMAScript 2015, and ECMAScript
2016.\footnote{The ECMAScript standard is now published yearly.} The role of the
specification is to make portability between different JavaScript
implementations possible by specifying the intended behavior of the various
language constructs and built-in functions of JavaScript. The specification also
makes it possible to develop formal semantics for the language, which can then
be used to, e.g., develop static analysis tools or guarantee that the invariants
required by the specification can be satisfied.

One such semantics is \lambdajs. It aims to capture
the semantics of JavaScript by defining a simple core language
(based on the untyped call-by-value lambda calculus), and then
transforming (or ``desugaring'') the full JavaScript language into this
simple core. The \lambdajs semantics was originally developed to model the 
ECMAScript 3 specification~\cite{Guha-al:ECOOP10}, and it was later
revised~\cite{Politz-al:DLS12} to handle ECMAScript 5.1
features.\footnote{The authors named the revised language S5; however,
we continue to name it \lambdajs for simplicity.}
The interpreter for the core language and the JavaScript-to-core
transformation were developed using 
OCaml~\cite{github:brownplt/LambdaS5};
the interpreter was tested using the official test262 test suite.
The \lambdajs semantics is useful for tool developers, as it
allows the tools to analyze JavaScript programs by working
on the simple core language of \lambdajs, having all the complexities
of JavaScript handled in the desugared code. It has already been
successfully used to verify ADsafe~\cite{Politz-al:SEC11}.

More recently, JSCert~\cite{Bodin-al:POPL14}, formalized the ECMAScript 5.1
specification in the form of a pretty-big-step
semantics~\cite{Chargueraud:ESOP13} expressed in the form of a Coq inductive
predicate. It aims to be as close to the (informal English) specification as
possible by maintaining close correspondence between the evaluation rules and
steps in the ES5 pseudocode. The project also includes JSRef, a reference
interpreter for JavaScript written in Coq and proven correct with respect to
JSCert. The interpreter is also tested using the official ECMA test suite,
test262. Thanks to the correctness proof, testing of JSRef increases the
confidence that JSCert faithfully models ES5.

The goal of this work is to formally relate the two semantics. Our motivation
is twofold. First, as the two semantics use a different approach, relating them
increases the confidence we can place in both of them, and has the side effect
of uncovering inconsistencies or other problems. Second, a formal and certified
version of \lambdajs, obtained by proving its correctness with respects to
JSCert, enables development of certified analysis tools using \lambdajs as the
core.

The main challenge lies in the complexities of the JavaScript language. The ES5
specification, upon which JSCert and \lambdajs are based, is 258 pages long and
has 16 chapters. The JSCert pretty-big-step semantics, which models the
specification, has over 900 rules; the entire Coq source code for the semantics
has over 10000 lines. The \lambdajs environment file, which contains the
ECMAScript algorithms expressed in core \lambdajs code, is almost 5000 lines
long.

As is done in \lambdajs, we focus on the \emph{strict mode} restriction of
JavaScript. Our contributions are as follows.
\begin{enumerate}
\item Two formalizations in Coq of \lambdajs: an inductive one, and an
  executable one. The formalizations coincide: they are shown to be correct and
  complete in relation to each other. This contribution is described in
  Section~\ref{sec:form-lambdajs}.
\item A formalization in Coq of desugaring from JavaScript to \lambdajs, as well
  as the initial \lambdajs environment in Coq. This contribution is described in
  Section~\ref{sec:desugaring}.
\item A correctness proof of the desugaring, stating that if a JavaScript
  program translated to \lambdajs successfully terminates, then the initial
  program has a semantics according to JSCert and the results match. This is
  described in Section~\ref{sec:corr-proof-lambdajs}.
\end{enumerate}

Note that we do not prove the converse, i.e., that if a JavaScript program has a
semantic in JSCert, then its \lambdajs translation successfully terminates and
the results match. The reason is technical and is described in
Section~\ref{sec:coq-techniques} alongside the techniques we implemented in Coq
for this development.

@@

\section{Formalization of \lambdajs}
\label{sec:form-lambdajs}

\asi{describe the changes to the language.}

@@

\section{Desugaring}
\label{sec:desugaring}

@@

\section{Correctness proof for \lambdajs desugaring}
\label{sec:corr-proof-lambdajs}

@@

\section{Coq techniques}
\label{sec:coq-techniques}

Proving the other direction requires doing an inversion of the full JSCert
semantics. Unfortunately, the techniques we have used do not scale to semantics
of that size.
